Implementation and comparison of register allocation to translate \lstinline!LLVM--! to x86

### 2023-12-30 article.tex
Jane approved

Compilation refers to the process of translating from one language to another, most often from a high-level programming language intended for humans to work with, to machine- or bytecode intended to be executed on a target architecture. This process can be divided into several distinct phases, which are grouped into one of two stages colloquially referred to as the \textit{frontend} and \textit{backend}, the former translating a high-level programming language to an \textit{intermediate representation} (IR) and the latter translating IR to executable machine code of a target architecture or bytecode of a target \textit{virtual machine} (VM).

@@

Although compilers can feasibly translate from the source language to machine code of the target platform directly, and doing so could even be more efficient, doing so hinders the portability as machine code targeting architecture \(A\) isn't necessarily useful for targeting architecture \(B\) further down the line.

Instead of translating directly from a source language to target machine code, most compiler frontends emit IR which is intended as an abstraction over particular CPU implementations. % FIXME: er det korrekt? lol
It is by no means executable on either \(A\) or \(B\) but is more often than not much closer to the abstract level of a CPU than the source language originally fed to the frontend.  This helps portability immensely, as a frontend emitting an IR no longer needs to specialize to a particular architecture, instead it can target as many as the backend supports.
Likewise, a compiler backend will support any frontend provided that they emit correct IR.

@@

Conversely, managed languages such as Java and C\# typically don't depend on the LLVM toolchain directly, research projects that attempt to compile associated VM bytecode to a native platform using LLVM (\cite{dotnet}). Meanwhile interpreted languages like JavaScript eagerly attempt use LLVM for \textit{just-in-time} (JIT) compilation % FIXME: this is not true at all lol

The modularization of the frontend and backend stages also help immensely.

@@

\subsubsection{Debugger Harness}
A more elaborate way of testing correctness of translation is to assert that there are no effects outside of those intended. One way of achieving this is to check if a group of x86 instructions representing a single LLVM-- operation change any values except those expected. An approach to this is to attach a debugger, insert a breakpoint before every LLVM-- instruction, and ensure that only the intended registers are altered.

To achieve this, only a few changes need to be made to the codebase. A debug flag is introduced, which inserts a label before every instruction by adding a \lstinline!Breakpoint! variant of the \lstinline!ins! type before the assembly emitted of that specific operation. In this a unique identifier as well as a list changed expected to be made is encoded. 
This includes a bitmask of GPRs that can possibly be changed (aside from the scrath registers of couse) as well as which parts of the stack may be modified. % TODO: implement the stack checker
In addition to this, each of the labels are output with a \lstinline!.globl! directive in the file header. A file is built with these extra symbols using the \lstinline!-d! flag.

@@

The above measures are significantly equivalent to one another, with all allocators (except greedy 0) seeming to converge towards being 37\% slower than Clang. This isn't particularly surprising, as the x86 translation is more concerned with preserving callee saved registers than necessary in pushing each and every one despite not being in use. Also, instead of pushing with the \lstinline!push! instruction it would be faster to simply subtract from the \lstinline!%rsp! register directly and using \lstinline!mov! directly.

What's more interesting and related to the allocation is the fact that greedy 0 is more than twice as slow. This is presumably because the \lstinline!fib! function that has 7 variables can assign each of them to a register for the first six allocators listed. However, the last one (greedy 0), has no assignable registers, so is forced to start spilling from the very beginning. To illusidate this hypothesis, here is the same benchmark but only for the greedy of decreasing \(k\) assignable registers:

@@

>>>>

Compilation refers to the process of translating from one language to another, most often from a high-level programming language intended for humans to work with, to machine- or bytecode intended to be executed on a target architecture. This process can be divided into several distinct phases, which are grouped into one of two stages colloquially referred to as the \textit{frontend} and \textit{backend}, The former is translating a high-level programming language to an \textit{intermediate representation} (IR) and the latter is translating IR to executable machine code of a target architecture or bytecode of a target \textit{virtual machine} (VM).

@@

Although compilers can feasibly translate from the source language to machine code of the target architecture directly, which could even be more efficient, doing so hinders the portability as machine code targeting architecture \(A\) isn't necessarily useful for targeting architecture \(B\) further down the line.

Instead of translating directly from a source language to target machine code, most compiler frontends emit IR which is intended as an abstraction over particular CPU implementations. % FIXME: er det korrekt? lol
It is by no means executable on either \(A\) or \(B\), but it is much closer to the instruction set executed on a CPU than the source language originally fed to the frontend.  This helps portability immensely, as a frontend emitting an IR no longer needs to specialize to a particular architecture, instead it can target as many as the backend supports.
Likewise, a compiler backend will support any frontend provided that they emit correct IR.

@@

Conversely, managed languages such as Java and C\# typically don't depend on the LLVM toolchain directly, some research projects attempt to compile associated VM bytecode to a native platform using LLVM (\cite{dotnet}).
%Meanwhile interpreted languages like JavaScript eagerly attempt use LLVM for \textit{just-in-time} (JIT) compilation % FIXME: this is not true at all lol

The modularization of the frontend and backend stages also helps immensely.

@@

\subsubsection{Debugger Harness}
A more elaborate way of testing correctness of translation is to assert that there are no effects except for those intended. One way of achieving this is to check if a group of x86 instructions representing a single LLVM-- operation change any values except those expected. An approach to this is to attach a debugger, insert a breakpoint before every LLVM-- instruction, and ensure that only the intended registers are altered.

To achieve this, only a few changes need to be made to the codebase. A debug flag is introduced, which inserts a label before every instruction by adding a \lstinline!Breakpoint! variant of the \lstinline!ins! type before the assembly emitted of that specific operation. A unique identifier as well as a list of changes expected to be made is encoded. 
This includes a bitmask of GPRs that can possibly be changed (aside from the scratch registers of course) as well as which parts of the stack may be modified. % TODO: implement the stack checker
In addition to this, each of the labels are output with a \lstinline!.globl! directive in the file header. A file is built with these extra symbols using the \lstinline!-d! flag.

@@

The measures in \ref{tab:dumb-fib}  are significantly equivalent to one another, with all allocators (except greedy 0) seeming to converge towards being 37\% slower than Clang. This isn't particularly surprising, as the x86 translation is more concerned with preserving callee saved registers than necessary in pushing each and every one despite not being in use. Also, instead of pushing with the \lstinline!push! instruction it would be faster to simply subtract from the \lstinline!%rsp! register directly and using \lstinline!mov! directly.

What's more interesting and related to the allocation is the fact that greedy 0 is more than twice as slow. This is presumably because the \lstinline!fib! function that has 7 variables can assign each of them to a register for the first six allocators listed. However, the last one (greedy 0), has no assignable registers, so it must necessarily start spilling from the very beginning. To elucidate this hypothesis, here is the same benchmark but only for the greedy of decreasing \(k\) assignable registers:

@@

\subsubsection{\lstinline!benches/fannkuch-redux.ll!}
\begin{table}[!ht]
\begin{center}
\begin{NiceTabular}{|c|cccccccc|}
\hline
arg & clang & simple 12 & simple 2 & briggs 12 & briggs 2 & linear & greedy 12 & greedy 0 \\
\hline
4 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 197975& \cellcolor[rgb]{0.75,0.9990234375,0.75} 0.93202x& \cellcolor[rgb]{0.771484375,0.9990234375,0.75} 1.08772x& \cellcolor[rgb]{0.75,0.9990234375,0.75} 0.94359x& \cellcolor[rgb]{0.84375,0.9990234375,0.75} 1.37648x& \cellcolor[rgb]{0.7900390625,0.9990234375,0.75} 1.16456x& \cellcolor[rgb]{0.771484375,0.9990234375,0.75} 1.09012x& \cellcolor[rgb]{0.7861328125,0.9990234375,0.75} 1.14835x\\
5 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 193800& \cellcolor[rgb]{0.7626953125,0.9990234375,0.75} 1.05122x& \cellcolor[rgb]{0.7744140625,0.9990234375,0.75} 1.10079x& \cellcolor[rgb]{0.7685546875,0.9990234375,0.75} 1.07640x& \cellcolor[rgb]{0.77734375,0.9990234375,0.75} 1.11231x& \cellcolor[rgb]{0.9296875,0.9990234375,0.75} 1.72450x& \cellcolor[rgb]{0.9453125,0.9990234375,0.75} 1.78685x& \cellcolor[rgb]{0.931640625,0.9990234375,0.75} 1.73329x\\
6 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 307184& \cellcolor[rgb]{0.8203125,0.9990234375,0.75} 1.28426x& \cellcolor[rgb]{0.8291015625,0.9990234375,0.75} 1.32022x& \cellcolor[rgb]{0.814453125,0.9990234375,0.75} 1.25973x& \cellcolor[rgb]{0.83203125,0.9990234375,0.75} 1.33286x& \cellcolor[rgb]{0.9990234375,0.9423828125,0.75} 3.81882x& \cellcolor[rgb]{0.9990234375,0.9404296875,0.75} 3.85798x& \cellcolor[rgb]{0.9990234375,0.939453125,0.75} 3.90989x\\
7 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 1312573& \cellcolor[rgb]{0.875,0.9990234375,0.75} 1.50533x& \cellcolor[rgb]{0.8896484375,0.9990234375,0.75} 1.56355x& \cellcolor[rgb]{0.8779296875,0.9990234375,0.75} 1.51619x& \cellcolor[rgb]{0.8828125,0.9990234375,0.75} 1.53672x& \cellcolor[rgb]{0.9990234375,0.861328125,0.75} 6.40226x& \cellcolor[rgb]{0.9990234375,0.8603515625,0.75} 6.44050x& \cellcolor[rgb]{0.9990234375,0.8642578125,0.75} 6.32867x\\
8 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 11252155& \cellcolor[rgb]{0.900390625,0.9990234375,0.75} 1.60713x& \cellcolor[rgb]{0.908203125,0.9990234375,0.75} 1.63817x& \cellcolor[rgb]{0.8818359375,0.9990234375,0.75} 1.52961x& \cellcolor[rgb]{0.90234375,0.9990234375,0.75} 1.61191x& \cellcolor[rgb]{0.9990234375,0.8349609375,0.75} 7.25974x& \cellcolor[rgb]{0.9990234375,0.8349609375,0.75} 7.25458x& \cellcolor[rgb]{0.9990234375,0.833984375,0.75} 7.28693x\\
9 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 115513112& \cellcolor[rgb]{0.8935546875,0.9990234375,0.75} 1.57771x& \cellcolor[rgb]{0.904296875,0.9990234375,0.75} 1.62155x& \cellcolor[rgb]{0.888671875,0.9990234375,0.75} 1.55882x& \cellcolor[rgb]{0.9990234375,0.98046875,0.75} 2.57223x& \cellcolor[rgb]{0.9990234375,0.82421875,0.75} 7.60550x& \cellcolor[rgb]{0.9990234375,0.8154296875,0.75} 7.87522x& \cellcolor[rgb]{0.9990234375,0.8212890625,0.75} 7.69289x\\
10 & \cellcolor[rgb]{0.75,0.9990234375,0.75} 1352747472& \cellcolor[rgb]{0.8857421875,0.9990234375,0.75} 1.54700x& \cellcolor[rgb]{0.935546875,0.9990234375,0.75} 1.74875x& \cellcolor[rgb]{0.89453125,0.9990234375,0.75} 1.58381x& \cellcolor[rgb]{0.9169921875,0.9990234375,0.75} 1.67440x& \cellcolor[rgb]{0.9990234375,0.818359375,0.75} 7.77269x& \cellcolor[rgb]{0.9990234375,0.8193359375,0.75} 7.74449x& \cellcolor[rgb]{0.9990234375,0.8173828125,0.75} 7.82170x\\
\hline
\end{NiceTabular}
\caption{\label{tab:fannkuch}Benchmark of \texttt{benches/sieven.ll} output by \mintinline{bash}{dune exec bench -- -f fib -n 1000}}
\end{center}
\end{table}
